Federated Learning's Blessing:\\ FedAvg has Linear Speedup

### 2021-07-04 source/app_nasgd_scvx_smooth.tex
Update on Overleaf.

\begin{thm}
	Let $\overline{\mathbf{v}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{v}_{T}^{k}$
	and set learning rates $\alpha_{t}=\frac{6}{\mu}\frac{1}{t+\gamma}$,  $\beta_{t-1}=\frac{3}{14(t+\gamma)(1-\frac{6}{t+\gamma})\max\{\mu,1\}}$. Then under Assumptions~\ref{ass:lsmooth},\ref{ass:stroncvx},\ref{ass:boundedvariance},\ref{ass:subgrad2} with full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right),
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}\sigma^{2}/\mu}{NT}+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right).
	\end{align*}
\end{thm}
%
%
\textbf{}%
\begin{comment}
This implies that $E$ canot be chosen $O(T^{\beta})$ for any $\beta>0$
without degrading the performance. This should be checked in experiments,
whether with partial participation if the communication round is set
to scale with $T$, the convergence deteriorates. This is in constrast
with the full participation case, where $E=O(\sqrt{\frac{T}{N}})$
is allowed. \textbf{If we can confirm that full participation allows
linear speedup with $\nu=N\cdot\max_{k}p_{k}\approx1$ and $E=O(\sqrt{\frac{T}{N}})$,
whereas partial participation only allows $E=O(1)$, then this would
an interesting phenomenon that is not reported by previous studies!}

>>>>

Recall that the Nesterov accelerated FedAvg follows the updates 
\begin{align*}
\mathbf{v}_{t+1}^{k} & =\mathbf{w}_{t}^{k}-\alpha_{t}\mathbf{g}_{t,k}, \hspace{1em}
\mathbf{w}_{t+1}^{k} =\begin{cases}
\mathbf{v}_{t+1}^{k}+\beta_{t}(\mathbf{v}_{t+1}^{k}-\mathbf{v}_{t}^{k}) & \text{if }t+1\notin\mathcal{I}_{E},\\
\sum_{k \in \cS_{t+1}}q_k\left[\mathbf{v}_{t+1}^{k}+\beta_{t}(\mathbf{v}_{t+1}^{k}-\mathbf{v}_{t}^{k})\right] & \text{if }t+1\in\mathcal{I}_{E}.
\end{cases}
\end{align*}

The proofs of convergence results for Nesterov Accelerated FedAvg consists of components that are direct analogues of the FedAvg case. We first state these analogue results before proving the main theorem. Like before, the proofs of the lemmas are deferred to after the main proof.

\begin{lemma}[\textbf{One step progress, Nesterov}] Let $\overline{\mathbf{v}}_{t}=\sum_{k=1}^{N}p_{k}\mathbf{v}_{t}^{k}$ in Nesterov accelerated FedAvg,
and suppose our functions satisfy Assumptions~\ref{ass:lsmooth},\ref{ass:stroncvx},\ref{ass:boundedvariance},\ref{ass:subgrad2}, and set step sizes $\alpha_{t}=\frac{6}{\mu}\frac{1}{t+\gamma}$,  $\beta_{t-1}=\frac{3}{14(t+\gamma)(1-\frac{6}{t+\gamma})\max\{\mu,1\}}$
	with $\gamma=\max\{32\kappa,E\}$ and $\kappa=\frac{L}{\mu}$, the updates of Nesterov accelerated FedAvg satisfy
\begin{align*}
\mathbb{E}\|\ov{v}_{t+1}-\vw^{\ast}\|^{2} & \leq\mathbb{E}(1-\mu\alpha_{t})(1+\beta_{t-1})^{2}\|\ov{v}_{t}-\vw^{\ast}\|^{2}+20E^{2}L\alpha_{t}^{3}G^{2}+(1-\alpha_{t}\mu)\beta_{t-1}^{2}\|(\ov{v}_{t-1}-\vw^{\ast})\|^{2}\\
& +\alpha_{t}^{2}\frac{1}{N}\nu_{\max}\sigma^{2}+2\beta_{t-1}(1+\beta_{t-1})(1-\alpha_{t}\mu)\|\ov{v}_{t}-\vw^{\ast}\|\cdot\|\ov{v}_{t-1}-\vw^{\ast}\|
\end{align*}
\label{lem:nest-scvxoner}
\end{lemma}

The one step progress result makes use of the same bound on the gradient variance in~Lemma~\ref{lem:bgv}, as well as a divergence bound analogous to Lemma~\ref{lem:bdw}, which we state below.
\begin{lemma}[\textbf{Bounding the divergence of $\vw_t^k$, Nesterov}]
Given Assumption~\ref{ass:subgrad2}, and assume that $\alpha_t$ is non-increasing, $\alpha_t \leq 2\alpha_{t+E}$, and $2\beta_{t-1}^{2}+2\alpha_{t}^{2}\leq1/2$ for all $t\geq 0$, $\overline{\mathbf{w}}_{t}=\sum_{k=1}^{N}p_{k}\mathbf{w}_{t}^{k}$ in Nesterov accelerated FedAvg satisfies
	\begin{align*}
	\mathbb{E}\left[\sum_{k=1}^{N}p_{k}\|\ov{w}_{t}-\vw_{t}^{k}\|^{2} \right]\leq16(E-1)^{2}\alpha_{t}^{2}G^{2}.
	\end{align*}
\label{lem:nest-bdw}
\end{lemma}

\begin{thm}
	Let $\overline{\mathbf{v}}_{T}=\sum_{k=1}^{N}p_{k}\mathbf{v}_{T}^{k}$ in Nesterov accelerated FedAvg
	and set learning rates $\alpha_{t}=\frac{6}{\mu}\frac{1}{t+\gamma}$,  $\beta_{t-1}=\frac{3}{14(t+\gamma)(1-\frac{6}{t+\gamma})\max\{\mu,1\}}$. Then under Assumptions~\ref{ass:lsmooth},\ref{ass:stroncvx},\ref{ass:boundedvariance},\ref{ass:subgrad2} with full device participation, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}\sigma^{2}/\mu}{NT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right),
	\end{align*}
	and with partial device participation with $K$ sampled devices at
	each communication round, 
	\begin{align*}
	\mathbb{E}F(\overline{\mathbf{v}}_{T})-F^{\ast}=\mathcal{O}\left(\frac{\kappa\nu_{\max}\sigma^{2}/\mu}{NT}+\frac{\kappa E^{2}G^{2}/\mu}{KT}+\frac{\kappa^{2}E^{2}G^{2}/\mu}{T^{2}}\right).
	\end{align*}
\end{thm}
%
%
\textbf{}%
\begin{comment}
This implies that $E$ canot be chosen $O(T^{\beta})$ for any $\beta>0$
without degrading the performance. This should be checked in experiments,
whether with partial participation if the communication round is set
to scale with $T$, the convergence deteriorates. This is in constrast
with the full participation case, where $E=O(\sqrt{\frac{T}{N}})$
is allowed. \textbf{If we can confirm that full participation allows
linear speedup with $\nu=N\cdot\max_{k}p_{k}\approx1$ and $E=O(\sqrt{\frac{T}{N}})$,
whereas partial participation only allows $E=O(1)$, then this would
an interesting phenomenon that is not reported by previous studies!}
